\documentclass[11pt]{article}
\usepackage{../../tex/math-cmds}
\usepackage{../../tex/analysis}
\usepackage{IEEEtrantools}
\usepackage{mathabx}
\usepackage{upgreek}


\title{Lecture Notes: Program Repair as Reachability}
\author{17-355/17-665/17-819: Program Analysis (Spring 2020)\\
  Claire Le Goues\footnote{These notes draw from Nguyen et al., ``Connecting
    Program Synthesis and Reachability: Automatic Program Repair using
    Test-Input Generation'', TACAS 2017 and
    a set of course materials generously provided by Wes Weimer.}\\
  {\tt clegoues@cs.cmu.edu}} 

\date{}

\begin{document}

\maketitle
There are useful correspondences between techniques we have used for
verification and those we then explored for synthesis and repair. For example,
verification condition generation began as a method to prove programs correct;
generating these conditions forwards rather than backwards allowed us to develop
a way to generalize testing through symbolic execution; we could then use
symbolic/concolic execution as a way to perform synthesis for repair of certain
types of defects.

Today, we will explore a formal connection between program synthesis (for
repair, specifically) and verification, formulated as a \emph{reachability}
problem.

\section{Template-based program synthesis for repair}

One general way to formulate program repair is as a problem of selecting and
appropriately instantiating one or more \emph{repair templates} at the
appropriate points in a program. We define a general syntax for a templated
program along the following lines (borrowing from the \WhileLang syntax, but
simplifying somewhat for the purposes of this discussion):

\newcommand\df{\bnfdef}
\newcommand\da{\bnfalt}
\newcommand\skips{\mbox{skip}}
\newcommand\ifs{\mbox{if}~ P ~\mbox{then}~ S_1 ~\mbox{else}~ S_2}
\newcommand\whiles{\mbox{while}~ P ~\mbox{do}~ S}

\[
\begin{array}{llllllllllllll}

S & \df & x := a    & a & \df & a_1 + a_2 \\
  & \da & \skips    &   & \da & a_1 - a_2 \\ 
  & \da & S_1;~ S_2 &   & \da & \fbox{$\displaystyle c_i $}  & \mbox{~~\emph{called a template parameter!}}\\
  & \da & \ifs      &   & \da & \ldots \\
  & \da & \whiles   &   & \\
\end{array}
\]


Given a templated program with template parameters $c_1 \ldots c_n$ and given
template values $\bar{v} = v_1 \ldots v_n$ (corresponding to expressions or
constants), we can \emph{instantiate} thatt emplate on those values to yield a
non-templated program. We can define instantiation in a straightforward,
syntax-directed way:

\begin{center}
\[
\begin{array}{lcl}
inst(\mbox{skip}, \bar{v}) & \rightarrow & \mbox{skip} \\
inst(S_1; S_2, \bar{v}) & \rightarrow & inst(S_1, \bar{v}); inst(S_2, \bar{v}) \\
inst(x := a, \bar{v}) & \rightarrow & x = inst(a, \bar{v}) \\
inst(\fbox{$\displaystyle c_i $}, \bar{v}) & \rightarrow & v_i 
\end{array}
\]
\end{center}

The \emph{template-based program synthesis problem} is then defined as follows: 
\begin{quote}
  Given a templated program $P$ with template parameters $c_1\ldots c_n$, and a
  set $T$ of input-output pairs (tests), do there exist template values
  $\bar{v}=v_1 \ldots v_n$ such that for all
  ${ \langle \alpha_0, \beta_0 \rangle \ldots \langle \alpha_n, \beta_n \rangle
  }$ in $T$, $(inst(P,\bar{v}))(\alpha_i) = \beta_i$?
\end{quote}

Note that we are ambivalent as to the mechanism used to identify $\bar{v}$, and
that many of the inductive techniques we have discussed either for synthesis
proper or for program repair specifically fit in this framing (consider
syntax-guided synthesis). 

We can extend this representation to program repair by constructing templated
programs from the original program and replacing potentially buggy lines with
potential template \fbox{$\displaystyle c_i $}. By synthesizing some code to
fill arbitrary hole, the repair effectively becomes ``delete buggy statement X
and replace with instantiated template Y.'' Overall, these templates
focus on expression-level manipulation; extending this framing to
statement-level modifications is an open problem. 

For the purposes of this exposition, we ignore fault localization, focus on
single-edit repairs, and on templates encoding linear combinations of variables.
Actual repair based on these ideas generalizes by trying several
locations/templates in some order. More complicated (e.g., non-linear) templates
are usable as well.

\section{Program Reachability}

The problem of reachability as applied to programs asks, very generally, whether
given a program $P$, a set of program variables $x_1 \ldots x_n$ and some
program label $L$, do there exist values $c_1 ... c_n$ such that $P$ with
$x_i = c_i$ reaches label $L$ in finite time?

We have seen this applied to finding bugs using symbolic execution (e.g.,
formulating buffer overflows as reaching an error state via program
transformation). Test generation can also be viewed as generating $c_i$ for test
inputs with $L$ corresponding to the end of a desired execution path. It is also
applicable to model checking, as we will see in future course lessons.

The following code example revisits the idea/intuition, calling back to our
prior discussions on test generation:

\begin{lstlisting}[language=C,basicstyle=\footnotesize,numbers=left]
int x, y; /* global input */
int P() { 
if (2 * x == y) {
  if (x > y + 10) 
     [ L ]
  return 0;
}
\end{lstlisting}

Here, $x = -20, y=-40$ reaches the label. 

\section{Reducing Synthesis To Reachability}

Both reachability and synthesis are undecidable in general. You may recall how
reductions work from prior theory or algorithms courses. In brief, Problem A is
reducible to Problem B if an efficient algorithm for B could be used as a
subroutine to solve A efficiently. A \emph{gadget} is a subset of a problem
instance that simulates the behavior of one of the fundamental units of a
different problem.


\subsection{Defining GadgetS2R}
Thus, given an instance of a synthesis (repair) problem, and assuming we have an
oracle that can solve reachability, let us convert the synthesis instance into a
reachability instance. If we can do this efficiently, any existing reachability
tool/technique (e.g., one that performs symbolic or concolic execution) could be
used to repair programs.

Give $Q$, a template program with a set of template parameters $S = \{ c_1,
\ldots, c_n\}$ and a set of finite tests $T = {(\alpha_1,\beta_1), \ldots}$,
  construct $GadgetS2R(Q,S,T)$ which returns a new program $P$ with a special
  location $L$, as follows:

\begin{enumerate}
\item For every template parameter $\bar{c_i}$, add a fresh global variable $v_i$. A
  solution to the reachability instance is an assignment of concrete values
  $c_i$ to variables $v_i$.
\item For every function $q \in Q$, define a similar function $q_{P} \in P$.
  The body of $q_P$ is the same as $q$, but with every reference to a template
  parameter $\bar{c_i}$ replaced with a reference to the corresponding new variable $v_i$
\item $P$ also contains a starting function $main_P$ that encodes the
  specification information from the test suite $T$ as a conjunctive expression
  $e$:

$e = \bigwedge\limits_{(\alpha_i,\beta_i) \in T} main_{QP}(\alpha_i) = \beta_i$

where $main_{QP}$ is a function in $P$ corresponding to the starting function
$main_Q$ in $Q$.  The body of $main_P$ is then a single conditional statement that
leads to a fresh target location $L$ iff $e$ is true.  
\item $P$ overall consists of the declaration of new variables, the functions
  $q_P$, and the starting function $main_P$
\end{enumerate}

\subsection{Illustrative example}

Consider the code sample we used in discussing semantics-based program repair:

\begin{lstlisting}[language=C,basicstyle=\footnotesize,numbers=left]
int is_upward(int in, int up, int down) {
  int bias, r;
  if (in)
    bias = down; // fix: bias = up + 100
  else
    bias = up;
  
  if (bias > down)
    r = 1;
  else
    r = 0;
  return r
  }
\end{lstlisting}

And a set of test cases that highlight the bug:

\begin{tabular}{l|rrr|rr|r}
\toprule
   & \multicolumn{3}{c}{\textbf{Inputs}} & \multicolumn{2}{c}{\textbf{Output}} &  \\
Test & in   & up  & down & expected & observed & Passed? \\\midrule
1    & 1 & 0 & 100 & 0 & 0 & Yes \\
2    & 1 & 11 & 110 & 1 & 0 & No \\
3    & 0 & 100 & 50 & 1 & 1 & Yes \\
4    & 1 & -20 & 60 & 1 & 0 & No \\
5    & 0 & 0 & -10 & 1 & 1 & Yes \\
\bottomrule
\end{tabular}

Assuming we select the buggy line and attempt a linear combination of variables
as a template, this transforms the buggy code into a templated program as so:


\begin{lstlisting}[language=C,basicstyle=\footnotesize,numbers=left]
int is_upward(int in, int up, int down) {
  int bias, r;
  if (in)
    bias = c0 + c1 * bias + c2 * in + c3 * up + c4 * down;
  else
    bias = up;
  
  if (bias > down)
    r = 1;
  else
    r = 0;
  return r;
  }
\end{lstlisting}


The reachability instance corresponding to this program then looks like:



\begin{lstlisting}[language=C,basicstyle=\footnotesize,numbers=left]
int c0, c1, c2, c3, c4; /* global input */

int P_is_upward(int in, int up, int down) {
  int bias, r;
  if (in)
    bias = c0 + c1 * bias + c2 * in + c3 * up + c4 * down;
  else
    bias = up;
  
  if (bias > down)
    r = 1;
  else
    r = 0;
  return r;
}
int main () {
  if (p_is_upward(1,0,100) == 0 &&
      p_is_upward(1,11,110) == 1 &&
      p_is_upward(0,100,50) == 1 &&
      p_is_upward(1,-20,60) == 1 &&
      p_is_upward(0,0,10) == 0 &&
      p_is_upward(0,0,-10) == 1) {
   [L] // label!
  } 
  return 0;
}
\end{lstlisting}

A valid solution to the reachability problem identifies global inputs \texttt{c0
  = 100, c1 = 0, c2 = 0, c3 = 1, c4 = 0}; this corresponds to instantiating the
template on line 4 with \texttt{bias = up + 100;}, a valid patch for this bug. 


\section{Proof of correctness}

To prove the correctness of this reduction, we must show that the constructed
reachability instance is solvable (with values $c_1 \ldots c_n$) iff the original
synthesis instance is solvable (with those same values). 
 %
The reachability instance is solved if those values cause execution to reach L,
while the synthesis instance is solved if those same values cause every test to
pass.

The proof uses standard operational semantics to reason about the meaning and
executions of programs. We use large-step semantics ($\Downarrow$) to reason
about synthesis, which focuses on the final value of the program (its behavior
on a test). W uses small-step semantics ($\rightarrow$) to reason about
reachability, where the intermediate steps matter (was a particular label
reached?). The proof uses induction on the structure of a derivation to show
that a property holds for all exeuctions of all programs, and weakest
preconditions to reason about the special conditional statements that encode
test cases.

The high-level proof structure is as follows:
\begin{itemize}
\item Lemma 1: The reachability instance method and the synthesis instance
  method agree on the values of all (non-template) variables.
\item Lemma 2: If the reachability instance reaches L from a state S (with
  values $c_1\ ldots c_n$), then that state and values model the weakest
  precondition of the synthesis instance method passing each test.
\item Theorem 1. The synthesis instance is solvable iff the reachability
  instance is solvable (with the same values). 
\end{itemize}

\vspace{1ex}

\subsection{Lemma 1: Agree on Vars}

The idea here is to show that the derived program ($p_q(\alpha_i)$) behaves the
same as the original program $q[c_0,\ldots,c_n](\alpha_i)$\footnote{$q[c_0,\ldots,c_n]$ is another way of saying ``q instantiated on $c_0,\ldots,c_n$''} when the new variables
$v_i$ in $P$ are assigned the values $c_i$. 

Formally, let $Q$ be the input synthesis instance method with template variables
$v_1 \ldots v_n$. Let $P = GadgetS2R(Q)$ be the reachability instance
corresponding to method $P$. For all states $E_1, E_2, E_3$, all values
$c_1, \ldots c_n$, all input values $x$, we seek to prove that it holds that:

\begin{quote}
  if $E_1(v_i) = c_i$, then $D_1 :: \langle P(x), E_1\rangle \Downarrow E_2$ iff \\
  $D_2 :: \langle inst(Q,\bar{c}), E_1 \rangle \Downarrow E_3$ \\
  and $\forall y \neq v_i, E_2(y) = E_3(y)$.
\end{quote}

The proof proceeds by induction on the structure of an operational semantics
derivation $D_1$. Let $E_1$ be arbitrary, except with $E_1(v_i) = c_i$.
Importantly, every $p_q(i) \in P$ corresponds to $q[c_1,\ldots,c_n](e) \in Q$,
in the sense that all subexpressions in $p_q$ and $q$ are identical
\emph{except} that references to template parameters \fbox{$\displaystyle c_i $} in $Q$
correspond to references to variables $v_i$ in $P$. Thus, by inversion, the
structure of $D_1$ corresponds exactly to the structure of $D_2$, except for the
variable references.  Thus, although for completeness we'd consider all cases
for the derivation $D_1$, in practice we'll only show the interesting case,
which is a read/assignment of a templated variable.  

\vspace{1ex}
\noindent \emph{Case:} Suppose $D_1$ (reachability instance) is:

 \[
\begin{array}{c}

\infer[\textit{assign}]
{ \langle a := v_i, E_1 \rangle \Downarrow E_2} {E_2 = E_1 [ a
  \mapsto E_1(v_i)] }
\end{array}
\]


\noindent By inversion and the construction of $P$, $D_2$ is: 

\[
\begin{array}{c}

\infer[\textit{assign}]
{ \langle a := exp, E_1 \rangle \Downarrow E_3} {E_3 = E_1 [ a
  \mapsto c_i] }
\end{array}
\]

\noindent 
where $exp = inst(\fbox{$\displaystyle c_i $}, \bar{c}) = c_i$

Now, we know that $E_2 = E_1 [ a \rightarrow E_1(v_i) ]$ and
$E_3 = E_1 [ a \rightarrow c_i ]$. We know that $E_2$ and $E_3$ agree on all
variables except $a$ (because they are the same as $E_1$, for all variables
except $a$). So, this obligation simplifies to showing that $E_1 (v_i) = c_i.$
This is actually one of the assumptions in the statement of the lemma.
(Intuitively, it means the reachability analysis assigned $c_i$ to each variable
$v_i$ to reach the label $L$.)

\vspace{1ex}
The other cases are direct, because they do not involve the template paramters
$v_i$: the inference rule used in $D_1$ will exactly mirror the inference rule
used in $D_2$.  

We have therefore established that the executions of $P$'s functions mirror $Q$'s
functions, modulo the template parameters, which are held constant.

\subsection{Lemma 2 (Reach L = Pass tests)}

We now establish that reaching $L$ in $P$ by assigning each $v_i$ the value
$c_i$ corresponds to $Q[c_i,\ldots,c_n]$ passing all of the tests. 

Let $Q$ be the input synthesis instance method with template variables
$v_1 \ldots. v_n$ and tests $\langle\alpha_1, \beta_n\rangle$ Let
$P = \mbox{GadgetS2R}(Q)$ be the reachability instance method \texttt{main}. 

We know that the execution of $P$ reaches $L$ starting from state $E_1$ iff
$E_1 \vDash wp(inst(Q, \bar{c})(\alpha_1), result = \beta_1) \wedge \ldots
wp(inst(Q,\bar{c}(\alpha_n), result = \beta_n)$ where $E_1(v_i) = c_i$.
%
By gadget construction, we also know there is only one label $L$ in $P$,
$\mathtt{if}~ e~ \mathtt{then}~[L]$, and that $e$ is of the form
$f(\alpha_1) = \beta_1 \wedge \ldots f(\alpha_n) = \beta_n$.

By standard weakest precondition definitions for if, conjunction, equality and
function calls, we have that $L$ is reachable iff
$E_1 \vDash wp(result := f(\alpha_1), result = \beta_1) \wedge \ldots wp(result
:= f(\alpha_n), result = \beta_n).$ What we \emph{want} is that $L$ is reachable
iff
$E_1 \vDash wp( result := inst(Q, \bar{c})(\alpha_1), result = \beta_1) \wedge
\ldots wp(result := inst(Q, c)(\alpha_n), result = \beta_n)$

So, we have to show that $E_1 \vDash wp(result := f(\alpha_i), result =
\beta_i)$ iff $E_1 \vDash wp(result := inst(Q, \bar{c})(\alpha_i), result =
\beta_i)$.  $f$ here is the method from GadgetS2R(Q).  

By the soundness and completeness of weakest preconditions wrt operational
semantics, we have $\langle result := f(\alpha_i), E_1\rangle \Downarrow E_2$
iff $E_2 \vDash result = \beta_i$.
%
By Lemma 1, we have $\langle result := inst(Q, \bar{c})(\alpha_i) ,
E_1 \rangle \Downarrow E_3$ iff $E_1(y)= E_3(y)$ for all $y \neq
v_i$.
%
Since ``result'' $\neq v_i$, $E_1(result) = E_3(result)$ (we know this from
lemma 1) and $E_3(result) = \beta_i$ (because of the weakest precondition
reasoning above). So, running the template program $Q$ instantiated with
$c_i = v_i$ on a test input produces the required output.

\subsection{Correctness}

This leads us to the correctness theorem, which is as follows: Let $Q$ be the
input synthesis instance method with template variables $v_1 ... v_n$ and tests
$\langle\alpha_1, \beta_n\rangle$. Let $P = GadgetS2R(Q)$ be the reachability
instance method \texttt{main}.

There exist parameter values $c_i$ such that for all
$\langle\alpha_i,\beta_i\rangle, inst(Q,\bar{c})(\alpha_i) = \beta_i$ iff there exists input
values $t_i$ s.t. the execution of $P$ with $v_i \rightarrow t_i$ reaches $L$.
The proof is from Lemma 2 with $t_i = c_i$. 

Note that we can also carry out a constructive reduction going in the other
direction. That is, suppose we are given an instance of program reachability.
\emph{Can we convert it into a program synthesis instance to solve
  it?}\footnote{We elide these details but note they are expanded upon in the
  original publication.}

\section{Implications}

Program reachability tools and techniques are much more mature than program
repair tools. This correspondence between the problems therefore suggests a way
to use reachability to attempt to fix bugs in programs. It proceeds by, for
every potentially buggy line, in some ranked order, and then for every possible
considered repair template, also in some ranked order, converting the repair
instance to a reachability instance and then calling an off-the-shelf
reachability tool (e.g., an SMT solver-based concolic execution engine like
KLEE). If the label is reachable, the discovered parameters in the satisfying
model can be returned for instantiation as a program patch.

Overall, this correspondance helps illustrate how all of the techniques we have
discussed (e.g., test generation, model checking, verification) are seeking, in
some way, to statically reason about dynamic execution. We can sometimes take
advantage of this correspondence to discover new techniques, as this particular
reduction demonstrates.


\bibliographystyle{abbrv}
\bibliography{repair}

\end{document}
